---
title: "descriptive statistics"
author: "Lucie Woellenstein"
output: html_document
---

This script produces:
  - a table summarising the descriptive statistics of a data set by outcome class
  - box plots of continuous variables
  - scatter plots of continuous variables versus outcome variables    

```{r setup, include=FALSE}
#set working directory of the whole document (includes 3 folders: raw_data, clean_data, research) - reset for every code chunk reading
knitr::opts_knit$set(root.dir="path/to/file") # put in correct path

# load libraries (these must be installed if not already)
library("tidyverse") 
library("rlist")
library("inspectdf")
library("WVPlots")

# create object for folder to write to 'research/report/' should be inside working directory
folder <- "/research/report/"
```

WRITE ANY FUNCTIONS REQUIRED IN SCRIPT 

```{r}
# function to get rounded numbers to nearest 5 & percentage by grouped var
get_prop_5 <- function(var, data) {
  data %>%
    group_by(.[var]) %>%
    summarize(NumPat = round(n()/5)*5,  # round count in group to the nearest 5 
              prop=round(sum(NumPat)/nrow(.)*100,2)) %>% # percentage based on the rounded NumPat
    # combine rounded count and percentage and add % sign to figure in column N
    mutate(N = paste0(as.character(NumPat), 
                      sep=" (",as.character(prop),sep="%)"),
           NumPat = NULL,
           prop = NULL)
}


# function to reshape proportion table 
# puts binary outcome classes into the two columns taking the N(%) for each predictor variable as their value
reshape_prop <- function(outcome, table, prefix) {
  outcome <-enquo(outcome) # put quotes around value so it can be read
  table %>%
    pivot_longer(c(-N,
                   -!!outcome),  # !! means to take quotes off, otherwise an error is thrown
                 names_to = "Variable",
                 values_to = "Value") %>%
    na.omit() %>%
    pivot_wider(names_from = !!outcome,
                names_prefix = prefix,
                values_from = N)  %>%
    filter(Value != "NO")
}

# function gets mean and standard deviation of continuous variables
get_mean_sd <- function(outcome, input, var, data) {
  outcome <- enquo(outcome) # puts quotes around value so it can be read
  var <- enquo(var)
  # get mean of variable per outcome class
  mean <- data %>% 
    filter(!!outcome==input) %>%  # !! takes quotes off again
    pull(!!var) %>% 
    mean() %>% 
    round(.,2)
  # calculate sd of variable per outcome class
  sd <- data %>% 
    filter(!!outcome==input) %>% 
    pull(!!var) %>% 
    sd() %>% 
    round(.,2)
  # paste together the mean and sd in ()
  output <- paste0(as.character(mean), 
                   sep=" (", as.character(sd), 
                   sep=")")
  output
}

# function reformats property variable categories to better names for table
reformat_PREVENT1 <- function(df) {
  df <- df %>%
    mutate(PROPERTY = as.factor(
    recode(
      PROPERTY,
      "accomm_tied_employment_armed_forces" = "Service tied accommodation",
      "family" = "Family",
      "hostel_B&B" = "Hostels & B&Bs",
      "roofless" = "Roofless",
      "friends_partners" = "Friends & partners",
      "home_ownership" = "Home ownership",
      "hospital" = "Hospital",
      "LA_tenancy" = "LA tenancy",
      "lodger" = "Lodger",
      "other" = "Other",
      "prison" = "Prison",
      "PRS_tenancy" = "PRS tenancy",
      "RSL_tenancy" = "RSL tenancy",
      "sofa_surfing" = "Sofa-surfing",
      "supported_accom" = "Supported accommodation",
      "unknown_refused" = "Unknown or refused"
    )
  ))
}

```

LOAD AND PREPARE DATA 

```{r}
# read data from /clean_data/ folder in working directory
data <- read_csv("/clean_data/clean_data.csv")

# prepare data 
# set variables to NULL that should be removed, change category names for binary variables, factorise categorical vars
data <- data %>%
  mutate(refused_vul = NULL, 
         time_error = NULL, 
         RSNSCT = as.factor(RSNSCT),
         ETHNICITY = as.factor(ethnicity),
         ethnicity = NULL,
         REASON = as.factor(REASON),
         PROPERTY = as.factor(PROPERTY),
         GENDER = as.factor(ifelse(gender==1, 
                                   "male", 
                                   "female")),
         gender = NULL,
         SUPPORT = as.factor(SUPPORT),
         other_activity = other,
         other = NULL,
         previous_approach = new_previous_approach,# give better name
         new_previous_approach = NULL,
         made_homelessness_application = as.factor(ifelse(made_homelessness_application == TRUE, 
                                                          "YES", 
                                                          "NO")),
         fin_vul = as.factor(ifelse(fin_vul == TRUE, 
                                    "YES",
                                    "NO")),
         phys_vul = as.factor(ifelse(phys_vul == TRUE, 
                                     "YES", 
                                     "NO")),
         mental_vul = as.factor(ifelse(mental_vul == TRUE, 
                                       "YES", 
                                       "NO")),
         unmetneed_vul = as.factor(ifelse(unmetneed_vul == TRUE, 
                                          "YES", 
                                          "NO")),
         lackofsupport_vul = as.factor(ifelse(lackofsupport_vul == TRUE, 
                                              "YES", 
                                              "NO")))

# check column types are as would be expected
data %>% 
  inspect_types %>% 
  show_plot()

# get rounded count and percentage n(%) for outcome variable - to know numbers in each class (required below)
get_prop_5("outcome", data)
```

CREATE DESCRIPTIVE STATISTICS TABLE
 - descriptive statistics by outcome class 
 - categorical variables include count per nearest 5 and percentage in each outcome class
 - continuous variables have mean and standard deviation per outcome class

```{r}
# make table of all categorical descriptive variables n(%) per outcome class
table <- rbind(
  get_prop_5(var = c("outcome", 
                     "GENDER"),
             data = data),
  get_prop_5(var = c("outcome",
                     "ETHNICITY"), 
             data = data),
  get_prop_5(var = c("outcome",
                     "PROPERTY"), 
             data = data),
  get_prop_5(var = c("outcome",
                     "REASON"), 
             data = data),
  get_prop_5(var = c("outcome",
                     "RSNSCT"), 
             data = data),
  get_prop_5(var = c("outcome",
                     "fin_vul"),
             data = data),
  get_prop_5(var = c("outcome",
                     "phys_vul"), 
             data = data),
  get_prop_5(var = c("outcome",
                     "mental_vul"),
             data = data),
  get_prop_5(var = c("outcome",
                     "unmetneed_vul"), 
             data = data),
  get_prop_5(var = c("outcome",
                     "lackofsupport_vul"), 
             data = data)
  ) %>%
  reshape_prop(outcome = outcome,
               ., 
               prefix = "Outcome_") # prefix should be the outcome variable name to appear in column header

# make table of all continuous vars mean(SD) per outcome class
cont_avgs <- data.frame(cbind(
  Variable = c("AGEATAPP (years)",  
               "APP2OUT (days)", 
               "ACTIVITY_COUNT", 
               "VULNERABILITY_COUNT"),
  Value = "mean (SD)",
  unsuccessful = c(get_mean_sd(outcome = outcome, 
                               input = "unsuccessful", # the "successful" outcome class
                               var = ageatapp, 
                               data = data),
                   get_mean_sd(outcome = outcome, 
                               input = "unsuccessful",
                               var = APP2OUT, 
                               data = data),
                   get_mean_sd(outcome = outcome, 
                               input = "unsuccessful",  # the "unsuccessful" outcome class
                               var = activity_count, 
                               data = data),
                   get_mean_sd(outcome = outcome, 
                               input = "unsuccessful",
                               var = vul_count, 
                               data = data)),
  successful = c(get_mean_sd(outcome = outcome,
                             input = "successful",
                              var = ageatapp, 
                              data = data),
                 get_mean_sd(outcome = outcome, 
                             input = "successful", 
                             var = APP2OUT, 
                             data =data),
                get_mean_sd(outcome = outcome, 
                            input = "successful",
                            var = activity_count,
                            data = data),
                 get_mean_sd(outcome = outcome, 
                             input = "successful", 
                             var = vul_count,
                             data = data))))

# bind together for a descriptive statistics table
table <- table %>%
  rbind(cont_avgs) %>%
  # rename columns "outcome variable - class (n=)"
  # fill in n= with stats from calculation in get_prop_5("outcome", data)
  rename("Outcome-unsuccessful (n=)" = unsuccessful) %>%
  rename("Outcome-successful (n=)" = successful) 

# view table to check it's good
print(table)  

# write table to research/report/ folder in working directory
write.csv(table, 
          file.path(folder, "descriptive_stats.csv"), 
          row.names = F)

```

CREATE BOX PLOTS OF TIME VARIABLES

```{r}
# pivot longer data set - the time variables that were used 
longer <- data %>%
  pivot_longer(c(APP2ACT, 
                 ACT2OUT, 
                 APP2OUT),
               names_to = "var",
               values_to = "value")

# set theme for plot (this one just changes the title)
mytheme <- theme(plot.title = element_text(size = 15, 
                                           hjust = 0.5))
# boxplots without outliers
p <- ggplot(longer, 
            aes(x = reorder(var, value), 
                y = value)) +
  geom_boxplot(outlier.shape = NA) +
  mytheme +
  # set the limits of the y-axis (should be)
  scale_y_continuous(limits = quantile(x = longer$value, 
                                       probs = c(0, 0.65))) + 
  ggtitle("Distribution of Time Variables (outliers removed)") +
  labs(y = "Time (days)",
       x = "Variable")

# view plot
p

ggsave(file.path(folder, "boxplots"),
      plot = p,
      device = png())
```

CREATING SCATTER PLOTS
 - for continuous variables and the outcome variable
 - first the outliers are found and removed

```{r}
# find Q1 Q3 IQR for variable in "successful" outcome class
summary(data %>% 
          filter(outcome == "successful") %>%
          select(APP2OUT))
# inner_upper1 <- # inner upper fence calculation: Q3 +1.5(IQR) 
# inner lower wasn't required here

# find Q1 Q2 IQR for variable in "unsuccessful" outcome class
summary(data %>% 
          filter(outcome == "unsuccessful") %>%
          select(APP2OUT))
# inner_upper2 <- # inner upper fence calculation: Q3 +1.5(IQR)
# inner lower wasn't required here

# remove minor and major outliers from data set for scatter plots in next section
data_out <- data %>%
  filter(outcome =="successful" & APP2OUT <= inner_upper1|
           outcome=="unsuccessful" & APP2OUT <= inner_upper2)
```

- now the Binary Y scatterplots are made for continuous variables versus outcome 

```{r}
# binary y scatter plots show the relationship between a continuous x independent variable and a binary y variable
scatter <- data_out %>%
  # must turn outcome variable to numeric or BinaryYScatterPlot won't work
  mutate(outcome = ifelse(outcome == "successful", 
                          1, 
                          0))


### MAKE SCATTER PLOTS 

# lapply was used here as there were two outcome variables in data set only one (outcome) included here
# lapply code retained for purposes of showing how to use it

# set y-axis name
yvar <- c("outcome")

# make list of the data frames of data without outliers and binary variable turned to numeric
frames <- list(scatter)

# plot scatter plots for all outcome variables
BYscatterplots <-
  lapply(1:length(titles), 
         function(x) BinaryYScatterPlot(frames[[x]],
                                        xvar = "APP2OUT",
                                        yvar = yvar[[x]],
                                        use_glm = TRUE)
         )

# name the graph in the list - more useful if multiple scatter plots were created
names(BYscatterplots) <- c("scatterplot1")

# view scatter plots
BYscatterplots

# save scatter plot
lapply(names(BYscatterplots), 
       function(x) ggsave(filename = file.path(folder, x),
                          plot = BYscatterplots[[x]],
                          device = png())
       )

# redo above with all continuous variables
```
