---
title: "Exploratory Data Analysis"
author: "Lucie Woellenstein"
output: pdf_document
---

Exploratory Data Analysis for any data sets analysed 
- relevant data sets and variable names (where applicable) must be changed where indicated 

```{r setup include=FALSE}
#set working directory of the whole document (includes 3 folders: raw_data, clean_data, research) - reset for every code chunk reading
knitr::opts_knit$set(root.dir="path/to/file") # put in correct path to working directory ('homelessness-prevention-single-adults-scotland')

# load necessary libraries (these must be installed if not already)
library("tidyverse") # loads dplyr, ggplot2, tidyr, purr, tibble, forcats
library("rlist")
library("inspectdf")
library("GGally")
library("skimr")
library("readr")
library("WVPlots")
# load in the bonferroni adjustment function 'chisq.post.hoc' from this webpage
source('https://raw.githubusercontent.com/PassionDrivenStatistics/R/master/ChiSquarePostHoc.R')

```

READ IN THE DATA 

``` {r}
# read in the data 
data <- read_csv("clean_data/clean_prevent1.csv", # put correct data set being explored should be in "your/working/directory/clean_data/"
                 col_types = cols(LACODE = col_factor(), # factorise any categorical variables
                                  outcome = col_factor(), 
                                  gender = col_factor(), 
                                  REASON = col_factor()))

# factorise categorical variables that were missed above
data <- data %>%
  mutate(PROPERTY = factor(PROPERTY), 
         ethnicity = factor(ethnicity))
```

EXAMINE DATA SET STRUCTURE USING INSPECT

```{r inspect}
# use inspect package to inspect the data 

# column types
data %>%
  inspect_types() %>%
  show_plot()


# inspect correlation coefficients of numeric columns
data %>%
  select(-person_id) %>% # remove person ID
  inspect_cor() %>%
  show_plot()

# inspect imbalance in categorical columns related to outcome var, each row should have 2 dots on graph
data %>%
  group_by(outcome) %>%
  inspect_imb() %>%
  show_plot()

# summarise categorical columns plot
data %>%
  inspect_cat() %>%
  show_plot()

# summarise categorical columns in table
data %>%
  inspect_cat()
```

EXAMINE ALL VARIABLES USING SKIMR PACKAGE

```{r skimr}
# use skimr package to explore data set further

# provides summary statistics for different variable types
skim(data)

# provides statistics on all variables crossed to with the outcome (e.g. missing, mean)
data %>%
  select(-person_id) %>% # remove irrelevant person_id
  group_by(outcome) %>%
  skim()
```

EXAMINE THE CONTINUOUS VARIABLES 
- histograms, density plots, summary statistics
- correlographs between all continuous variables

```{r}
# summary statistics for continuous variables
summary(data$APP2ACT) 
summary(data$ACT2OUT)
summary(data$APP2OUT)
summary(data$ageatapp)

# build histogram of continuous variables
hist(data$APP2ACT) 
hist(data$ACT2OUT)
hist(data$APP2OUT)
hist(data$ageatapp)

# density graphs - check distributions of continuous variables
plot(density(data$APP2ACT)) 
plot(density(data$ACT2OUT))
plot(density(data$ageatapp))

# plot correlation between continuous variables using a correlogram with ggally
# possible to do this with binary variables
data %>%
  select(ageatapp, 
         APP2ACT,
         ACT2OUT, 
         APP2OUT) %>%
  ggpairs(title = "Correlogram for numerical variables")
```


EXAMINE CORRELATION
 - measures the strength and direction of a linear association between two variables
 - continuous versus continuous variables (use pearson correlation)
 - continuous versus binary predictor variables (use point biseral - i.e. "pearson")
 - continuous versus binary outcome variable (use point biseral - i.e. "pearson" method)
 -if any ordinal variables present, use KENDALL correlation)

```{r}

# POINT BISERAL CORRELATION - PEARSON CORRELATION
# correlation between continuous variables and binary outcome data 
cbind(
  outcome = data$outcome,
  homelessapp = data$made_homelessness_application,
  ageatapp = data$ageatapp,
  APP2OUT = data$APP2OUT,
  APP2ACT = data$APP2ACT,
  vulcount = data$vul_count,
  actcount = data$activity_count
) %>%
  cor(method = "pearson")

# correlation between binary outcome variables and binary predictor variables (vulnerability and activity variables)
# correlation between binary predictor variables (vulnerability and activity variables)
cbind(
  outcome = data$outcome,
  homelessapp = data$made_homelessness_application,
  finvul = data$fin_vul,
  mentalvul = data$mental_vul,
  physvul = data$phys_vul,
  lackvul = data$lackofsupport_vul,
  unmetvul = data$unmetneed_vul,
  financial = data$financial,
  welfare = data$welfare_benefits,
  mediation = data$mediation,
  miscellaneous = data$miscellaneous,
  rights = data$homelessness_rights,
  general = data$general_housing_advice_info_assistance,
  renting = data$renting,
  owning = data$owning_buying,
  health_emp_support = data$health_employment_adaptations_support_services,
  type1 = data$activity_type_1,
  type2 = data$activity_type_2,
  type3 = data$activity_type_3,
  by1 = data$activity_by_1,
  by2 = data$activity_by_2,
  by3 = data$activity_by_3
) %>%
  cor(method = "pearson")

```

CREATE SCATTER PLOTS
- graphs relationship between continuous and binary outcome variable

```{r}
# set titles of each graph you want to make
titles <- c("age vs outcome", "APP2OUT vs outcome", "APP2ACT vs outcome")

# set x-axis variable for each xvar in tiles above (must be in same order)
xvar <- c("ageatapp", "APP2OUT", "APP2ACT")

# set y-axis variable - the outcome variable
yvar <- "outcome"
BYscatterplots <- lapply(1:length(titles), 
                         function(x) BinaryYScatterPlot(data, 
                                                        xvar = xvar[[x]], 
                                                        yvar = yvar,
                                                        title = titles[[x]]))
print(BYscatterplots)
# dev.off() # if plot doesn't show when printing try this
```

EXAMINE THE CATEGORICAL VARIABLES
- cross tabulations of each categorical predictor variable with binary outcome variable
- cross tabulations of each categorical predictor variable with each other

```{r}

# table for just one variable - numbers in each category
# do for all variables if desired
table(data$financial_assistance_advice)

# plot categorical variables against each other (bar graphs)
plot(data$gender, data$ethnicity)
plot(data$financial_assistance_advice, data$REASON)

### CROSS TABULATIONS - clues for correlation between variables ###


# cross tabulations two categorical variables - actual numbers
# how many of each category cross over between two categorical variables?
data$gender %>%
  table(data$ethnicity) %>%
  addmargins()

data$PROPERTY %>%
  table(data$REASON) %>%
  addmargins()

# Cross tabulation of two categorical variables - proportions instead
# measure for all pairs of categorical variables
# helps spot collinearity
as.data.frame(data$financial_assistance_advice %>% # becomes the variable in first column
                table(data$REASON) %>% # becomes variable in second column
                prop.table(margin = 2) %>% # margin = 2 => proportion of row category (financial assistance) in the column category (reason)
                round(digits = 2)*100
              )

### create cross tabulations of each categorical variable with the outcome variable to be modelled later on ###

# create vector of all categorical variables in data set
cat_vars <- c("PROPERTY", 
              "outcome", 
              "gender", 
              "REASON", 
              "ageatapp", 
              "LACODE", 
              "ethnicity", 
              "fin_vul", 
              "phys_vul", 
              "mental_vul", 
              "unmetneed_vul", 
              "lackofsupport_vul", 
              "refused_vul", 
              "new_previous_approach", 
              "made_homelessness_application", 
              "time_error", 
              "activity_type_1", 
              "activity_type_2", 
              "activity_type_3",
              "financial_assistance_advice", 
              "general_housing_advice_tenancy_rights_advice", 
              "client_informed_of_rights_under_homelessness_legislation", 
              "rent_repairs_referrals_neogtiation_with_landlords", 
              "other", 
              "mediation_outreach", 
              "referral_to_health_social_work_employment_services", 
              "tenancy_social_care_support_adaptations_to_property",
              "mortgage_homeownership_advice", 
              "help_to_remain_in_accommodation", 
              "activity_by_1", 
              "activity_by_3", 
              "activity_by_2", 
              "SUPPORT")

# use lapply function to create cross tabulations of each variable in cat_vars and the outcome variable
outcome_tables <- lapply(data[, cat_vars], 
                         function(x) round(100 * prop.table(table(data$outcome, # repeat and replace with each variable in cat_vars
                                                                  x, # each of the variables in cat_vars
                                                                  exclude = NA), 
                                                            margin = 2), 
                                           digits = 1))
outcome_tables


# repeat above process to make cross tabulations between each of the non binary categorical predictor variables 

```

EXAMINE CORRELATION BETWEEN CATEGORICAL VARIABLES

- use Chi Square test and Bonferroni test (function written in bonferroni.r called in set up chunk)
- Chi square test for independence checking. This will let you know if there is a correlation, but not the direction or strength
- Bonferroni test lets you check in the correlation is between every category of a variable (i.e. which categories are significant)
- Larger numbers in the Chi square test mean no relationship 
- Smaller numbers means there is a relationship
- P-value indicates the probability of getting the observed counts assuming that the two variables are not related (counts under null hypothesis)
- the smaller the chi square p-value the more surprising it would be to get the observed counts if the null hypothesis was true (i.e. smaller the p-vaue, more likely the two are associated with each other)

```{r error=FALSE}

# (significance <0.05)

# map the chisq.test to the cross tabulations created above
outcome_X2 <- map(outcome_tables, chisq.test)
chisq.test(outcome_tables$ethnicity)

print(outcome_X2)

# repeat the chisq.test with cross tabulations between non binary categorical predictor variables

# BONFERRONI TEST

# usual accepted p value
p <- 0.05 

### Bonferroni adjustment p values for non binary categorical variables


# number of tests run - the number of pairs between the categories of two variables that showed correlation in chi square test
num_tests <- c(45, 66, 106) # adjust accordingly to number of pairs

# adjusted p value for that number of categories in a variable
adjp <- c(
  "adjp9", 
  "adjp12", 
  "adjp16"
  ) 

#  adjusted p-values to compare against in the bonferroni test below
adjp <- cbind(
  adjp, 
  p / num_tests
  )

# perform bonferroni test

# extract the observed values from the chi.test above
obs <- lapply(
  outcome_X2, # replace with each individual chi square test above (e.g. between two categorical predictor variables)
  function(x) x$observed
  )

# extract the observations for each categorical variable that showed correlation with another categorical variable (here compared with outcome) in the chi square test above
observations <- list(
  reason = obs$REASON, 
  property = obs$PROPERTY, 
  ethnicity = obs$ethnicity
  )


# perform the bonferroni test on each of the observations pulled out above
outcome_bonferroni_tests <- lapply(
  observations, 
  # chisq.post.hoc comes from the bonferroni.r function written and called in during set up
  function(x) chisq.post.hoc(x, 
                             control = "bonferroni", 
                             popsInRows = FALSE)[, 1:2]
  )


# if the raw.p value (bonferroni significance) printed below is higher than the adjusted p-value for that number of pairs (adj values above),
# the relationship between those two pairs are not significant
print(outcome_bonferroni_tests)
```
