---
title: "LR graphs"
author: "Lucie Woellenstein"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    df_print: paged
    highlight: haddock
    code_folding: hide
---

## Introduction

This R markdown script is used to graph performance metrics of each of the Logistic Regression (LR)
models built in "/research/analysis/prediction_models.Rmd".

These LR models were built with the maximum relevance minimum redundancy feature selection methods, 
meaning as many models were built as there were predictor variables (K / features) in the data set. 
Each model had one additional variable in it. 

The example data that can be used to run this script is found in "/research/visualise/data/". 

This first chunk loads the packages that we will need
```{r}
# load libraries needed
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(janitor))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(socviz))
suppressPackageStartupMessages(library(patchwork))
```

The next chunk builds custom functions required to build the graph 

```{r}
# write custom function to graph all performance metrics for each K number of features in the LR model
LR_graph <- function(df, title, best) {
  df %>%
    ggplot(mapping = aes(x = K, y = value, group = variable)) +
    theme_minimal() +
    scale_color_manual(values = c("red", "blue", "dark green", "hot pink", "yellow")) +
   geom_line(aes(color = variable)) + 
   labs(title = title,
        y = "Metrics") +
    theme(plot.title = element_text(size=14)) +
    scale_x_continuous(name = "K: Number of Features in Model",
                       minor_breaks = seq(1, 35, 1)) +
    scale_y_continuous(limits = c(0,1),
                       breaks = c(0,.20,.40,.60,.80,.100)) +
    guides(col = guide_legend("Performance Metric")) + 
    geom_vline(xintercept = best, 
               color = "grey", 
               linetype = "longdash") +
    theme(legend.position = "none") 
}
```

Load in the data

```{r}
# read in the csv of LR performance metrics. 
# the current script must be in the same folder where data is found (/research/visualise/)
df <- read_csv("/data/LR_metrics.csv")

# manipulate the data into long form for graphing
df <- df %>%
  dplyr::select(K, F1, Accuracy, AUROC, Sensitivity, Specificity) %>%
  gather(key = "variable", 
         value = "value", 
         - K)
```


```{r}
# use custom function to graph metics, remember to put in the optimal k for the dashed line
graph <- LR_graph(df = df, 
                  title= "Glasgow", 
                  best = 23)

# view the graph 
graph
```

```{r}
# save the graph in the directory folder chosen 
ggsave("/homelessness-prevention-single-adults-scotland/LR_metrics.png", 
       plot = graph, 
       width = 12, 
       height = 9,
       bg="white")
```
