---
title: "OR plot"
author: "Lucie Woellenstein"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    df_print: paged
    highlight: haddock
    code_folding: hide
---
## Introduction

This rmarkdown script produces Odds Ratio plots. 

The odds ratios (ORs) came from the Logistic Regression (LR) and Elastic Net (ENet) models produced by  '/research/analysis/prediction_models.rmd'. 

The OR plot only includes ORs that had significant p-values in the LR model and were included 
in both the LR and ENet models. 

The example data that can be used to run this script is in "/research/visualise/data/".
This data holds the ORs for mdoels classifying successful and unsuccessful Housing Options outcomes 
in the Glasgow PREVENT1 data set.

The following chunk sets the baseline characteristics for the plot

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(janitor))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(socviz))
suppressPackageStartupMessages(library(patchwork))


#Baseline plot settings
theme_set(theme_minimal(base_family = "Helvetica", base_size = 24) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  axis.line.x = element_line(colour = 'black', size=0.5,
                                             linetype='solid'),
                  axis.line.y = element_line(colour = 'black', size=0.5,
                                             linetype='solid'),
                  strip.text = element_text(face = "bold"),
                  plot.caption = element_text(colour = "#AAAAAA"),
                  plot.margin = margin(3,15,3,3,"mm")))
options(digits = 2, scipen = 999)

`%nin%` <- negate(`%in%`)
```

## Load and clean data

```{r}
#Import
or_data <- read_excel("data/ORs.xlsm", # this excel file has LR and ENet coefficients and ORs. Data folder is inside visualise folder
                      sheet = "OR graph") %>% # this is the correct tab on the excel file with just the ORs to include in the graph
  #assign true NA values
  mutate(across(Enet:upper, ~na_if(.x, "NA"))) %>% 
  #Coerce from character to numeric variables
  mutate(across(Enet:upper, as.numeric)) %>% 
  #round to 2 decimal places (function from package socviz)
  round_df()

or_data
```


## Plots

Create distinct plots for each binary variable to be "patched" together later and continuous variables and categorical variables

```{r}
#quick look at filtered values
or_data %>% 
  filter(str_detect(Term, "ETHNICITY|SUPPORT|REASON|PROPERTY|RSNSCT"))  %>% 
 #arrange in order of ENet/LR Odds Ratio
  arrange(LR) %>%
  na.omit()
  #We will factorise the Term variable according to this order
  #*IMPORTANT*
  #The reference should be the first term in the factor to make the plot easier to understand.
```

```{r, fig.width=16, fig.height=14}
or_data %>% 
  filter(str_detect(Term, "PROPERTY|SUPPORT|REASON")) %>% 
  na.omit() %>%
  add_row(., Term = "SUPPORT No", Enet = 1, LR = 1, upper = 1, lower = 1) %>% 
  add_row(., Term = "REASON Applicant terminated accommodation", Enet = 1, LR = 1, upper = 1, lower = 1) %>% 
  add_row(., Term = "PROPERTY Employment/armed\nforces accommodation", Enet = 1, LR = 1, upper = 1, lower = 1) %>% 
  mutate(Term = factor(Term,
                       levels = c("PROPERTY Employment/armed\nforces accommodation",
                                  "PROPERTY Hostels & B&Bs",
                                  "PROPERTY Roofless",
                                  "PROPERTY Prison",
                                  "PROPERTY Sofa-surfing",
                                  "PROPERTY Hospital",
                                  "PROPERTY LA tenancy",
                                  "PROPERTY Home ownership",
                                  "PROPERTY RSL tenancy",
                                  "REASON Applicant terminated accommodation",
                                  "REASON Eviction UK borders",
                                  "REASON Service tied accommodation",
                                  "REASON Violent abusive dispute",
                                  "REASON Fleeing",
                                  "REASON Overcrowding",
                                  "REASON HO advice",
                                  "REASON Financial / benefits issues",
                                  "SUPPORT No",
                                  "SUPPORT Yes",
                                  "SUPPORT Not known"))) %>% 
  #now start  the plot
  ggplot(aes(fct_rev(Term), LR, ymin=lower, ymax = upper)) +
  #LR values
  geom_pointrange(aes(color = "Logistic Regression")) +
  #Add the Enet values
  geom_point(aes(y = Enet, colour = "Enet"), size = 2.5) +
  #Solid line at y=1 - aids interpretation
  geom_hline(yintercept = 1, size = 1.2, colour = "black") +
  # Log the y axis - you might need to play around with the limits
  scale_y_log10(limits = c(0.02, 8),
                           breaks = c(0.05, 0.1, 0.25, 0.5, 0.75,
                                      2, 4, 8)) +
  #We'll add a custom legend
  scale_colour_manual(name = "Model", 
                      breaks = c("Enet", "Logistic Regression"),
                      values = c("Enet" = "#CC6677", 
                                 "Logistic Regression" = "#4477AA")) +
  theme(legend.position = "none") +
  coord_flip() +
  labs(x = "",
       y = "", 
       #subtitle handy for the combined plot
       subtitle = "Categorical variables") -> categorical_plot
categorical_plot
```

### CONTINUOUS variables

```{r}
#quick look at filtered values
or_data %>% 
  filter(Term %in% c("ACTIVITY COUNT", "AGEATAPP", "APP2OUT", "VULNERABILITY COUNT")) %>% 
 #arrange in order of ENet/LR Odds Ratio
  arrange(LR) %>%
  na.omit()
  #We will factorise the Term variable according to this order
  #*IMPORTANT*
  #I'm guessing that "Referral services" is the reference value here, 
  # I may be wrong - you can sort this out. The reference should be the
  # first term in the factor to make the plot easier to understand.
```


```{r, fig.width=16, fig.height=4}
or_data %>% 
  filter(Term %in% c("ACTIVITY COUNT", "AGEATAPP", "APP2OUT"))  %>% 
  na.omit() %>%
  mutate(Term = factor(Term,
                       levels = c("ACTIVITY COUNT",
                                  "AGEATAPP", 
                                  "APP2OUT"))) %>% 
  #now start  the plot
  ggplot(aes(fct_rev(Term), LR, ymin=lower, ymax = upper)) +
  #LR values
  geom_pointrange(aes(colour = "Logistic Regression")) +
  #Add the Enet values
  geom_point(aes(y = Enet, colour = "Enet"), size = 2.5) +
  #Solid line at y=1 - aids interpretation
  geom_hline(yintercept = 1, size = 1.2, colour = "black") +
  # Log the y axis - you might need to play around with the limits
  scale_y_log10(limits = c(0.02, 8),
                           breaks = c(0.05, 0.1, 0.25, 0.5, 0.75,
                                      2, 4, 8)) +
  #We'll add a custom legend
  scale_colour_manual(name = "Model", 
                      breaks = c("Enet", "Logistic Regression"),
                      values = c("Enet" = "#CC6677", 
                                 "Logistic Regression" = "#4477AA")) +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "",
       y = "", 
       #subtitle handy for the combined plot
       subtitle = "Continuous variables") -> continuous_plot
continuous_plot
```
Binary variables (VULNERABILITY, GENDER, ALL ACTIVITY RELATED VARS)

### BINARY VULNERABILITY variables

```{r}
or_data %>% 
  filter(str_detect(Term, c("ACTIVITY|VULNERABILITY|GENDER|PREVIOUS APPROACH")), 
         Term != "ACTIVITY COUNT", 
         Term != "VULNERABILITY COUNT") %>%
  na.omit() %>%
  arrange(LR) 
```

```{r, fig.width=16, fig.height=10}
or_data %>% 
  filter(str_detect(Term, c("ACTIVITY|VULNERABILITY|GENDER|PREVIOUS APPROACH")), 
         Term != "ACTIVITY COUNT", 
         Term != "VULNERABILITY COUNT")   %>% 
  na.omit() %>% 
  mutate(Term = factor(Term,
                       levels = c(
                                  "ACTIVITY Help moving",
                                  "ACTIVITY Homelessness rights",
                                  "ACTIVITY HO advice",
                                  "ACTIVITY Financial assistance",
                                  "ACTIVITY Help remaining",
                                  "ACTIVITY Other",
                                  "ACTIVITY BY 3 Other",
                                  "ACTIVITY TYPE 2 Casework",
                                  "VULNERABILITY Lack of support",
                                  "VULNERABILITY Mental health"))) %>% 
  #now start  the plot
  ggplot(aes(fct_rev(Term), LR, ymin=lower, ymax = upper)) +
  #LR values
  geom_pointrange(aes(color = "Logistic Regression")) +
  #Add the Enet values
  geom_point(aes(y = Enet, colour = "Enet"), size = 2.5) +
  #Solid line at y=1 - aids interpretation
  geom_hline(yintercept = 1, size = 1.2, colour = "black") +
  # Log the y axis - you might need to play around with the limits
  scale_y_log10(limits = c(0.02, 8),
                           breaks = c(0.05, 0.1, 0.25, 0.5, 0.75,
                                      2, 4, 8)) +
  #We'll add a custom legend
  scale_colour_manual(name = "Model", 
                      breaks = c("Enet", "Logistic Regression"),
                      values = c("Enet" = "#CC6677", 
                                 "Logistic Regression" = "#4477AA")) +
  theme(legend.position = "none") +
  coord_flip() +
  labs(x = "",
       y = "", 
       #subtitle handy for the combined plot
       subtitle = "Binary variables") -> binary_plot
binary_plot
```

## Combine plots

```{r, fig.width=18, fig.height=24}
patch_design <- c(area(1, 1, 2, 4),
                  area(3, 1, 14, 4),          # Specify design of grid
                  area(15, 1, 20, 4))

combined <- continuous_plot/categorical_plot/binary_plot + 
  labs(y = "Odds Ratio 95% CI (log scale)",
       x = "") +
 # plot_annotation(title = "Odds ratios for Significant Terms in both General Linear Models") +
  plot_layout(guides="collect", design = patch_design) &
  theme(legend.position = "bottom")

combined
```

```{r, eval=FALSE}
# save the image in the report folder
ggsave("homelessness-prevention-single-adults-scotland/research/report/or_plot.png", 
       combined, 
       width = 15, 
       height = 22, 
       bg = "white")
```

