---
title: "sankey diagrams"
author: "Lucie Woellenstein"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    df_print: paged
    highlight: haddock
    code_folding: hide
---

This script creates a Sankey Diagram showing the flow from original 
property accommodation types of HO approaches to the final outcome property accommodation type.
This sankey diagram was built on the raw_data (unavailable here due to data sensitivity)
and clean_data (fake synthetic data example under clean_data folder in main)

SET UP 

```{r setup, include=FALSE}
#set working directory of the whole document (includes 3 folders: raw_data, clean_data, research) - reset for every code chunk reading
knitr::opts_knit$set(root.dir="path/to/file") # put in correct path to working directory ('homelessness-prevention-single-adults-scotland')

# load necessary libraries (these must be installed if not already)
library("tidyverse")
library("rlist")
library("inspectdf")
library("networkD3")
library("htmlwidgets")
library("data.table")

# create object for where you will write to
folder <- "/research/report/"
```

WRITE NECESSARY FUNCTIONS

```{r}

# get percentages rounded to the nearest 5 percentage in each category)
# can take multiple variables and creates a cross tabulation with proportions
get_round_prop <- function(var, data) {
  data %>%
    group_by(.[var]) %>%
    summarize(NumPat = round(n()/5)*5, 
              prop=round(sum(NumPat)/nrow(.)*100,2)) %>%
    ungroup()
}

# function renames category values using mutate() ad recode() for PROPERTY variable
# for visualisation purposes
reformat_PREVENT1 <- function(df) {
  df <- df %>%
    mutate(PROPERTY = as.factor(
    recode(
      PROPERTY,
      "accomm_tied_employment_armed_forces" = "Service tied accommodation",
      "family" = "Family",
      "hostel_B&B" = "Hostels & B&Bs",
      "roofless" = "Roofless",
      "friends_partners" = "Friends & partners",
      "home_ownership" = "Home ownership",
      "hospital" = "Hospital",
      "LA_tenancy" = "LA tenancy",
      "lodger" = "Lodger",
      "other" = "Other",
      "prison" = "Prison",
      "PRS_tenancy" = "PRS tenancy",
      "RSL_tenancy" = "RSL tenancy",
      "sofa_surfing" = "Sofa-surfing",
      "supported_accom" = "Supported accommodation",
      "unknown_refused" = "Unknown or refused"
    )
  ))
}
  
```


LOAD CLEANED DATA 

```{r}
data <- read_csv("/clean_data/clean_data.csv")

# clean the data (remove unnecessary variables, rename binary variables where necessary, factorise categorical variables)
data <- data %>%
  mutate(refused_vul = NULL, 
         time_error = NULL, 
         RSNSCT = as.factor(RSNSCT),
         ETHNICITY = as.factor(ethnicity),
         ethnicity = NULL,
         REASON = as.factor(REASON),
         PROPERTY = as.factor(PROPERTY),
         GENDER = as.factor(ifelse(gender==1, 
                                   "male", 
                                   "female")),
         gender = NULL,
         SUPPORT = as.factor(SUPPORT),
         other_activity = other,
         other = NULL,
         previous_approach = new_previous_approach,# give better name
         new_previous_approach = NULL,
         made_homelessness_application = as.factor(ifelse(made_homelessness_application == TRUE, 
                                                          "YES", 
                                                          "NO")),
         fin_vul = as.factor(ifelse(fin_vul == TRUE, 
                                    "YES",
                                    "NO")),
         phys_vul = as.factor(ifelse(phys_vul == TRUE, 
                                     "YES",
                                     "NO")),
         mental_vul = as.factor(ifelse(mental_vul == TRUE, 
                                       "YES", 
                                       "NO")),
         unmetneed_vul = as.factor(ifelse(unmetneed_vul == TRUE, 
                                          "YES", 
                                          "NO")),
         lackofsupport_vul = as.factor(ifelse(lackofsupport_vul == TRUE, 
                                              "YES",
                                              "NO")))

# check column types is what you expect
data %>% 
  inspect_types %>% 
  show_plot()
```

READ IN RAW UNCLEANED DATA
- retrieves the outcome accommodation locations that were pooled into the binary outcome variable
- pools the categories of  the raw outcome as the property variable was pooled when data was cleaned

```{r}

# read in raw data change the unique id name get just the outcome and person id and remove duplicates
raw_data <- read_csv("/raw_data/raw_data.csv")

# rename person id column so it has no spaces
names(raw_data)[1] <- 'person_id'

# removes any duplicates (i.e. where an approach had more than one column)
raw_outcome <- raw_data %>%
  mutate(raw_outcome = OUTCOME, 
         OUTCOME = NULL) %>%
  select(person_id, 
         raw_outcome) %>%
  group_by(person_id, 
           raw_outcome) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  filter(rank == 1)

# Pool raw_outcome (which holds outcome accommodation) into the same categories that the PROPERTY variable was
raw_outcome <- raw_outcome %>%
  mutate(raw_outcome = case_when(
    raw_outcome %in% c(1,24) ~ " LA tenancy",
    raw_outcome %in% c(5,26) ~ " PRS assured tenancy",
    raw_outcome %in% c(6,7,8,17) ~ " Hostels or womens refuge",
    raw_outcome == 98 ~ " Remained in accommodation",
    raw_outcome == 11 ~ " Family & friends",
    raw_outcome %in% c(12,27) ~" Other tenancy or lodger",
    raw_outcome == 13 ~ " Other known",
    raw_outcome == 14 ~ " Not known",
    raw_outcome %in% c(15,25) ~ " RSL tenancy",
    raw_outcome == 16 ~ " PRS short assured tenancy",
    raw_outcome %in% c(18,22,23) ~ " Supported accomm & nursing home",
    raw_outcome %in% c(19,20,21) ~ " Home ownership",
    raw_outcome == 28 ~ " Prison",
    raw_outcome == 29 ~ " Hospital",
    raw_outcome == 30 ~ " Lost contact",
    raw_outcome == 99 ~ " Made HL1 application")) %>%
  mutate(raw_outcome = as.factor(raw_outcome))


```

PREPARE ONE DATA FRAME FOR INPUT INTO SANKEY DIAGRAM

```{r}
# add the raw_outcome variable to the cleaned data set 
data <- data %>% 
  left_join(raw_outcome, by = "person_id") %>%
  mutate(rank = NULL)

# use reformat_PREVENT1 function written above to rename property categories for better visualisation
data <- reformat_PREVENT1(data)

# turn data frame to data table (to mutate raw_outcome values in next step)
data <- as.data.table(data)

# where property and raw_outcome were the same, change raw_outcome to remained in accommodation
data[PROPERTY == "Friends & partners" & raw_outcome == " Family & friends", 
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "Family" & raw_outcome == " Family & friends", 
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "Home ownership" & raw_outcome == " Home ownership",
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "Hospital" & raw_outcome == " Hospital", 
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "Hostels & B&Bs" & raw_outcome == " Hostels or womens refuge",
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "LA tenancy" & raw_outcome == " LA tenancy", 
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "Prison" & raw_outcome == " Prison", 
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "RSL tenancy" & raw_outcome == " RSL tenancy", 
     raw_outcome:= " Remained in accommodation"]

data[PROPERTY == "Supported accommodation" & raw_outcome == " Supported accomm & nursing home", 
     raw_outcome:= " Remained in accommodation"]

# turn data table back into data frame for sankey diagram
data <- as.data.frame(data)

# see the percentage (to nearest 5) in each category or cross tabulation of each category and arrange by size
as.data.frame(get_round_prop("PROPERTY", 
                             data)) %>%
  filter(NumPat != 0) %>%
  arrange(NumPat)


as.data.frame(get_round_prop("raw_outcome", 
                             data))  %>%
  filter(NumPat != 0) %>%
  arrange(NumPat)

as.data.frame(get_round_prop(c("PROPERTY", 
                               "raw_outcome"), 
                             data))  %>%
  filter(NumPat != 0) %>%
  arrange(NumPat)

```

MAKE SANKEY DIAGRAM
- requires the creation of links to travel along (property to raw_outcome)
- requires the creation of nodes on either end of the link (values of property and values of raw_outcome)

```{r sankey diagrams}
# create links from property variable to the raw_outcome variable 
links <- as.data.frame(
  get_round_prop(
    c("PROPERTY", 
      "raw_outcome"), 
    data)) %>%
  mutate(
    source = PROPERTY,
    target = raw_outcome,
    PROPERTY = NULL,
    raw_outcome = NULL
  ) %>%
  relocate(
    prop,  
    .after = last_col()       
    )

# take out any links that had no one travelling along it
links <- links %>%
  filter(prop != 0)

print(links)

# make the nodes (starting node: property) & end node: raw_outcome)
nodes <- data.frame(
  name = c(
  as.character(links$source),
  as.character(links$target)
  ) %>%
  unique())

print(nodes)

# ensure that the IDsource and IDtarget of links has the correct names 
links$IDsource <- match(links$source,
                        nodes$name) - 1
links$IDtarget <- match(links$target, 
                        nodes$name) - 1

# create the sankey diagram
sankey <- sankeyNetwork(Links = links, 
                        Nodes = nodes,
                        Source = "IDsource", 
                        Target = "IDtarget",
                        Value = "prop",
                        NodeID = "name",
                        fontSize = 15)

# find percentages for counts in each node
Perc <- links %>%
  group_by(source) %>%
  mutate(sPerc = paste0(round(sum(NumPat) / sum(links$NumPat) * 100, 2), "%")) %>%
  group_by(target) %>%
  mutate(dPerc = paste0(round(sum(NumPat) / sum(links$NumPat) * 100, 2), "%")) %>%
  pivot_longer(c(target, source))%>%
  mutate(Perc = ifelse(name == "target",
                       dPerc, 
                       sPerc)) %>%
  select(NumPat, 
         value, 
         Perc) %>%
  group_by(value, 
           Perc) %>%
  summarise(NumPat = sum(NumPat))
  

print(Perc)

# add the percentages to the nodes in diagram
sankey$x$nodes <- right_join(sankey$x$nodes, 
                             Perc, 
                             by = c("name" = "value"))
print(sankey$x$nodes)


sankey <- htmlwidgets::onRender(sankey, '
                      function(el,x) {
                        d3.select(el).selectAll(".node text")
                          .text(d => d.name + " " + d.Perc)
                      }')

# view the sankey diagram
sankey

# save the sankey widget as an html
saveWidget(sankey, 
           file=paste0(folder,"sankey.html"))
```

