---
title: "VIM Plots"
author: "Lucie Woellenstein"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    df_print: paged
    highlight: haddock
    code_folding: hide
---

## Introduction

This R markdown script graph the top 20 Variable Importance Measures (VIMs) in the models built in /research/analysis/prediction_models.Rmd). 

Example data that can be used to make this graph is available under research/visualise/data/VIMs/. 
The example are the VIMs from the models classifying successful and unsuccessful Housing Options outcomes
in the Glasgow PREVENT1 sample. 

The raw and cleaned data that the models were built on is not made available in this repository 
due to data sensitivity / confidentiality reasons. 
A fake synthetic example of the data set is available in /clean_data/clean_data.csv.

The following chunk loads the libraries necessary. Install these using install.packages("package name").

```{r setup, include=FALSE}
# set working directory for the whole notebook (this should be the highest level of the github repo 'homelessness-prevention-single-adults-scotland')
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/homelessness-prevention-single-adults-scotland")  # amend as necessary to where your repo sits

# load packages
library("tidyverse")
library("gridExtra")
library("grid")
```

Write the functions required in the script

```{r}
# create function to make a graph with the top 20 VIMs for a model
# VIMdf is a data frame containing all VIMs for a model produced in research/analysis/prediction_models.Rmd
plotvim <- function(VIM_df, title) {
  VIM_df %>% 
    mutate(rowname = forcats::fct_inorder(rowname)) %>%
    head(20) %>%
    ggplot() +
    geom_col(aes(x = rowname, y = Overall),
             fill = "grey",
             alpha = 0.9) +
    coord_flip()  +
    geom_text(
      aes(
        x = rowname,
        y = Overall,
        label = sprintf("%0.1f", round(Overall, 1)) # sprintf ensures 1 decimal point
      ),
      hjust = "inward",
      colour = "black"
    ) +
    labs(title = title,
         x = "Term",
         y = "Scaled Variable Importance Measure") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.minor = element_blank())
}

# function to rename the categories of variables for better visualisation in graph
reformat_PREVENT1 <- function(VIM_df) {
  VIM_df <- VIM_df %>%
    mutate(rowname = as.factor(
      recode(
        rowname,
        "PROPERTY4"	=	"PROPERTY Service tied accommodation",
        "PROPERTY6"	=	"PROPERTY Family"	,
        "PROPERTY10"	=	"PROPERTY Hostels & B&Bs",
        "PROPERTY11"	=	"PROPERTY Roofless"	,
        "PROPERTY6"	=	"PROPERTY Friends & partners",
        "PROPERTY5"	=	"PROPERTY Home ownership",
        "PROPERTY8"	=	"PROPERTY Hospital",
        "PROPERTY1"	=	"PROPERTY LA tenancy",
        "PROPERTY14"	=	"PROPERTY Lodger",
        "PROPERTY13"	=	"PROPERTY Other",
        "PROPERTY7"	=	"PROPERTY Prison",
        "PROPERTY3"	=	"PROPERTY PRS tenancy",
        "PROPERTY2"	=	"PROPERTY RSL tenancy",
        "PROPERTY12"	=	"PROPERTY Sofa-surfing",
        "PROPERTY9"	=	"PROPERTY Supported accommodation",
        "PROPERTY15"	=	"PROPERTY Unknown or refused",
        "REASON13"	=	"REASON Asked to leave"	,
        "REASON5"	=	"REASON Discharge"	,
        "REASON9"	=	"REASON Non-violent dispute"	,
        "REASON8"	=	"REASON Violent abusive dispute"	,
        "REASON17"	=	"REASON Eviction UK borders"	,
        "REASON10"	=	"REASON Fleeing"	,
        "REASON15"	=	"REASON HO advice"	,
        "REASON11"	=	"REASON Harassment"	,
        "REASON16"	=	"REASON Financial / benefits issues"	,
        "REASON2"	=	"REASON Landlord action"	,
        "REASON14"	=	"REASON Other leaving accommodaation"	,
        "REASON7"	=	"REASON Other loss accommodation"	,
        "REASON12"	=	"REASON Overcrowding"	,
        "REASON6"	=	"REASON Condition unsuitable"	,
        "REASON4"	=	"REASON Service tied accom"	,
        "REASON1"	=	"REASON Rent arrears / default"	,
        "REASON3"	=	"REASON Applicant termination"	,
        "ETHNICITY11"	=	"ETHNICITY White Scottish"	,
        "ETHNICITY9"	=	"ETHNICITY White British" 	,
        "ETHNICITY7"	=	"ETHNICITY Other White"	,
        "ETHNICITY12"	=	"ETHNICITY White Polish"	,
        "ETHNICITY2"	=	"ETHNICITY Asian"	,
        "ETHNICITY6"	=	"ETHNICITY Other"	,
        "ETHNICITY5"	=	"ETHNICITY Not known"	,
        "ETHNICITY3"	=	"ETHNICITY Caribbean / Black"	,
        "ETHNICITY4"	=	"ETHNICITY Mixed"	,
        "ETHNICITY8"	=	"ETHNCITIY Refused",
        "ACTIVITY_TYPE21"	=	"ACTIVITY TYPE 2"		,
        "ACTIVITY_TYPE11"	=	"ACTIVITY TYPE 1"		,
        "ACTIVITY_TYPE31"	=	"ACTIVITY TYPE 3"		,
        "ACTIVITY_BY21"	=	"ACTIVITY BY 2"	,
        "ACTIVITY_BY11"	=	"ACTIVITY BY 1"	,
        "ACTIVITY_BY31"	=	"ACTIVITY BY 3"	,
        "RSNSCT2"	=	"RSNSCT Need to leave accommodation"	,
        "RSNSCT3"	=	"RSNSCT Property unavailable in future"	,
        "MENTAL_VUL1"	=	"VULNERABILITY Mental health"	,
        "LACKOFSUPPORT_VUL1"	=	"VULNERABILITY Lack of support"	,
        "FIN_VUL1"	=	"VULNERABILITY Financial"	,
        "PHYS_VUL1"	=	"VULNERABILITY Physical"	,
        "UNMETNEED_VUL1"	=	"VULNERABILITY Unmet need"	,
        "REFUSED_VUL1" = "VULNERABILITY refused",
        "GENDER1" = "GENDER Female",
        "PREVIOUS_APPROACH1" = "Previous approach",
        "ACTIVITY_FINANCIAL_ASSISTANCE1" = "ACTIVITY_FINANCIAL_ASSISTANCE",
        "ACTIVITY_HELP_MOVING1" = "ACTIVITY_HELP_MOVING",
        "ACTIVITY_HELP_REMAINING1" = "ACTIVITY_HELP_REMAINING",
        "ACTIVITY_HO_ADVICE1" = "ACTIVITY_HO_ADVICE",
        "ACTIVITY_HOME_OWNERHSIP_ADVICE1" = "ACTIVITY_HOME_OWNERHSIP_ADVICE",
        "ACTIVITY_HOMELESSNESS_RIGHTS1" = "ACTIVITY_HOMELESSNESS_RIGHTS",
        "ACTIVITY_LANDLORD_NEGOTIATIONS1" = "ACTIVITY_LANDLORD_NEGOTIATIONS",
        "ACTIVITY_MEDIATION1" = "ACTIVITY_MEDIATION",
        "ACTIVITY_OTHER1" = "ACTIVITY_OTHER",
        "ACTIVITY_REFERRAL_SERVICES1" = "ACTIVITY_REFERRAL_SERVICES",
        "ACTIVITY_SOCIAL_CARE_ADAPTIONS1" = "ACTIVITY_SOCIAL_CARE_ADAPTIONS",
      )
    ))
}

# pulls out the top 5 most important variables
pull_top <- function(VIMdf, n) {
  VIMdf %>%
    mutate(Variable = forcats::fct_inorder(rowname),
           rowname = NULL) %>%
    head(5) %>% 
    pull(Variable) %>%
    as.character()
}

# set the destination filepath - should be in working directory then "/research/report/"
destfilepath <- "research/report"
```

Load in the data 

```{r}
# set the working directory for this chunk where the data is found 
setwd("/homelessness-prevention-single-adults-scotland/research/visualise/data/VIMs") # amend as necessary to where your repo sits
folder <- list.files(pattern="*.csv")

# read the VIMs
VIMdfs <- lapply(folder, function(x) read_csv(x))

# use this naming convention for the graphs
names(VIMdfs) <- c("AdaboostM1", "Bagged Adaboost", "Elastic Net", "Logistic Regression", "Random Forest", "Extreme Gradient Boosting")

# rename the categories using the custom function built above
VIMdfs <- lapply(VIMdfs, reformat_PREVENT1)

# view the VIM data frames
VIMdfs
```

Graph the individual VIMs

```{r}
# graph the top 20 VIMs for each algorithm putting the full algorithm name in each chart title
VIM_graphs <- lapply(1:length(VIMdfs), 
                     function(x) 
                       plotvim(VIMdfs[[x]], 
                               title = paste(names(VIMdfs[x]))))

# name the individual graphs in the list of graphs - used for saving them out
names(VIM_graphs) <- c("Ada", "BAda", "ENet", "LR", "RF", "XGB")

# view the graphs
VIM_graphs

# save each individual graph 
lapply(names(VIM_graphs),
       function(x)
         ggsave(
           filename = file.path(destfilepath, 
                                paste(x, 
                                      "_VIM.png",
                                      sep = "")),
           plot = VIM_graphs[[x]],
           width = 5,
           height = 3.5
         ))
```

