---
title: "VIM Table"
author: "Lucie Woellenstein"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    df_print: paged
    highlight: haddock
    code_folding: hide
---

## Introduction 

This R markdown script makes a table of all Variable Importance Measures (VIMs) that appeared in any of the
models' top five most important variables (models built in /research/analysis/prediction_models.Rmd). 

Example data that can be used to make this graph is available under research/visualise/data/VIMs/. 
The example are the VIMs from the models classifying successful and unsuccessful Housing Options outcomes
in the Glasgow PREVENT1 sample. 

The raw and cleaned data that the models were built on is not made available in this repository due to data sensitivity / confidentiality reasons. 

A fake synthetic example of the data set is available in /clean_data/clean_data.csv.

The following chunk loads the libraries necessary. Install these using install.packages("package name").

```{r setup, include=FALSE}
# set working directory for the whole notebook (this should be the highest level of the github repo 'homelessness-prevention-single-adults-scotland')
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/homelessness-prevention-single-adults-scotland") # put in your path

# load packages
library("tidyverse")
library("gridExtra")
library("grid")
```

The following chunk builds any functions required later on. 
```{r}
# rename category names to better names for visualisation
# this is specific to the data that is loaded in the next chunk 
# function to rename the categories of variables for better visualisation in graph
reformat_PREVENT1 <- function(VIM_df) {
  VIM_df <- VIM_df %>%
    mutate(rowname = as.factor(
      recode(
        rowname,
        "PROPERTY4"	=	"PROPERTY Service tied accommodation",
        "PROPERTY6"	=	"PROPERTY Family"	,
        "PROPERTY10"	=	"PROPERTY Hostels & B&Bs",
        "PROPERTY11"	=	"PROPERTY Roofless"	,
        "PROPERTY6"	=	"PROPERTY Friends & partners",
        "PROPERTY5"	=	"PROPERTY Home ownership",
        "PROPERTY8"	=	"PROPERTY Hospital",
        "PROPERTY1"	=	"PROPERTY LA tenancy",
        "PROPERTY14"	=	"PROPERTY Lodger",
        "PROPERTY13"	=	"PROPERTY Other",
        "PROPERTY7"	=	"PROPERTY Prison",
        "PROPERTY3"	=	"PROPERTY PRS tenancy",
        "PROPERTY2"	=	"PROPERTY RSL tenancy",
        "PROPERTY12"	=	"PROPERTY Sofa-surfing",
        "PROPERTY9"	=	"PROPERTY Supported accommodation",
        "PROPERTY15"	=	"PROPERTY Unknown or refused",
        "REASON13"	=	"REASON Asked to leave"	,
        "REASON5"	=	"REASON Discharge"	,
        "REASON9"	=	"REASON Non-violent dispute"	,
        "REASON8"	=	"REASON Violent abusive dispute"	,
        "REASON17"	=	"REASON Eviction UK borders"	,
        "REASON10"	=	"REASON Fleeing"	,
        "REASON15"	=	"REASON HO advice"	,
        "REASON11"	=	"REASON Harassment"	,
        "REASON16"	=	"REASON Financial / benefits issues"	,
        "REASON2"	=	"REASON Landlord action"	,
        "REASON14"	=	"REASON Other leaving accommodaation"	,
        "REASON7"	=	"REASON Other loss accommodation"	,
        "REASON12"	=	"REASON Overcrowding"	,
        "REASON6"	=	"REASON Condition unsuitable"	,
        "REASON4"	=	"REASON Service tied accom"	,
        "REASON1"	=	"REASON Rent arrears / default"	,
        "REASON3"	=	"REASON Applicant termination"	,
        "ETHNICITY11"	=	"ETHNICITY White Scottish"	,
        "ETHNICITY9"	=	"ETHNICITY White British" 	,
        "ETHNICITY7"	=	"ETHNICITY Other White"	,
        "ETHNICITY12"	=	"ETHNICITY White Polish"	,
        "ETHNICITY2"	=	"ETHNICITY Asian"	,
        "ETHNICITY6"	=	"ETHNICITY Other"	,
        "ETHNICITY5"	=	"ETHNICITY Not known"	,
        "ETHNICITY3"	=	"ETHNICITY Caribbean / Black"	,
        "ETHNICITY4"	=	"ETHNICITY Mixed"	,
        "ETHNICITY8"	=	"ETHNCITIY Refused",
        "ACTIVITY_TYPE21"	=	"ACTIVITY TYPE 2"		,
        "ACTIVITY_TYPE11"	=	"ACTIVITY TYPE 1"		,
        "ACTIVITY_TYPE31"	=	"ACTIVITY TYPE 3"		,
        "ACTIVITY_BY21"	=	"ACTIVITY BY 2"	,
        "ACTIVITY_BY11"	=	"ACTIVITY BY 1"	,
        "ACTIVITY_BY31"	=	"ACTIVITY BY 3"	,
        "RSNSCT2"	=	"RSNSCT Need to leave accommodation"	,
        "RSNSCT3"	=	"RSNSCT Property unavailable in future"	,
        "MENTAL_VUL1"	=	"VULNERABILITY Mental health"	,
        "LACKOFSUPPORT_VUL1"	=	"VULNERABILITY Lack of support"	,
        "FIN_VUL1"	=	"VULNERABILITY Financial"	,
        "PHYS_VUL1"	=	"VULNERABILITY Physical"	,
        "UNMETNEED_VUL1"	=	"VULNERABILITY Unmet need"	,
        "REFUSED_VUL1" = "VULNERABILITY refused",
        "GENDER1" = "GENDER Female",
        "PREVIOUS_APPROACH1" = "Previous approach",
        "ACTIVITY_FINANCIAL_ASSISTANCE1" = "ACTIVITY_FINANCIAL_ASSISTANCE",
        "ACTIVITY_HELP_MOVING1" = "ACTIVITY_HELP_MOVING",
        "ACTIVITY_HELP_REMAINING1" = "ACTIVITY_HELP_REMAINING",
        "ACTIVITY_HO_ADVICE1" = "ACTIVITY_HO_ADVICE",
        "ACTIVITY_HOME_OWNERHSIP_ADVICE1" = "ACTIVITY_HOME_OWNERHSIP_ADVICE",
        "ACTIVITY_HOMELESSNESS_RIGHTS1" = "ACTIVITY_HOMELESSNESS_RIGHTS",
        "ACTIVITY_LANDLORD_NEGOTIATIONS1" = "ACTIVITY_LANDLORD_NEGOTIATIONS",
        "ACTIVITY_MEDIATION1" = "ACTIVITY_MEDIATION",
        "ACTIVITY_OTHER1" = "ACTIVITY_OTHER",
        "ACTIVITY_REFERRAL_SERVICES1" = "ACTIVITY_REFERRAL_SERVICES",
        "ACTIVITY_SOCIAL_CARE_ADAPTIONS1" = "ACTIVITY_SOCIAL_CARE_ADAPTIONS",
      )
    ))
}

# function pulls out the top 5 most important variables in each model
pull_top <- function(VIMdf, n) {
  VIMdf %>%
    mutate(Variable = forcats::fct_inorder(rowname),
           rowname = NULL) %>%
    head(5) %>% 
    pull(Variable) %>%
    as.character()
}
destfilepath <- "research/report"
```

Load in the data

```{r}
# set working directory for this chunk only (put in your path)
setwd("/homelessness-prevention-single-adults-scotland/research/visualise/data/VIMs")
# list the csv files in the folder
# ensure that the folder only includes csv that are needed for this script
folder <- list.files(pattern="*.csv")
# read all csv into a data frame
VIMdfs <- lapply(folder, 
                 function(x) 
                   read_csv(x))

# use this naming convention for the top 5 table 
names(VIMdfs) <- c("AdaboostM1", "Bag-Adaboost", "ENet", "LR", "RF", "XGB")
VIMdfs

# reformat the data frames to change category names using custom function above
VIMdfs <- lapply(VIMdfs, 
                 reformat_PREVENT1)
# view the VIM data frames
VIMdfs
```


```{r top 5 table}
# apply custom pull_top function to all VIMdfs
# makes distinct list of all of them VIMs in any of the models' top five most important variables
top5 <- lapply(VIMdfs, 
               pull_top, 
               n = 5)

# turn list into a vector
top5 <- c(top5[[1]], top5[[2]], top5[[3]], top5[[4]], top5[[5]], top5[[6]])

# turn to a dataframe table
table <- as.data.frame(table(top5))

# name the columns
names(table) <- c("Variable", "Appearances")

# add VIMs to the data frame
table <- table %>% 
  left_join(., VIMdfs$ENet, by=c("Variable"="rowname")) %>% 
  left_join(., VIMdfs$LR, by=c("Variable"="rowname")) %>%
  left_join(.,VIMdfs$AdaboostM1, by=c("Variable"="rowname")) %>%
  left_join(., VIMdfs$`Bag-Adaboost`, by=c("Variable"="rowname")) %>% 
  left_join(., VIMdfs$RF, by=c("Variable"="rowname")) %>% 
  left_join(., VIMdfs$XGB, by=c("Variable"="rowname")) 

# name columns with the additional columns that have been added
names(table) <- c("Variable", "Appearances", "ENet", "LR mRMR", "AdaboostM1", "Bag-Adaboost", "RF", "XGB")

# round the VIMs to one decimal point
table <- table %>% 
  mutate_at(vars(-c(Variable, Appearances)), funs(round(.,1)))

# order by number of appearances
table <- table %>%
  arrange(desc(Appearances))

# view the table
table

# save the csv of the table
write_csv(table, 
          file = file.path(destfilepath, "top_5_VIMs_table.csv"))
```
