---
title: "VIM Plots"
author: "Lucie Woellenstein"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    df_print: paged
    highlight: haddock
    code_folding: hide
---

## Introduction

This R markdown script graphs the Variable Importance Measures (VIMs) for all the models that were created in /research/analysis/prediction_models.Rmd. As with the other visualisation scripts (or_plot, sankey_diagrams, descriptive statistics), this example was run on the VIMs of the models run on the Glasgow PREVENT1 data set. This data set is not made available due to data sensitivity / confidentiality reasons. A fake synthetic example of the data set is available here /clean_data/clean_data.csv.

The following chunk loads the libraries necessary. Install these using install.packages("package name").

```{r}
library("tidyverse")
library("gridExtra")
library("grid")
```

Write the functions required in the script

```{r}
# create function to make a graph with the top 20 VIMs for a model
# VIMdf is a data frame containing all VIMs for a model produced in research/analysis/prediction_models.Rmd
plotvim <- function(VIM_df, title) {
  VIM_df %>% 
    mutate(rowname = forcats::fct_inorder(rowname)) %>%
    head(20) %>%
    ggplot() +
    geom_col(aes(x = rowname, y = Overall),
             fill = "grey",
             alpha = 0.9) +
    coord_flip()  +
    geom_text(
      aes(
        x = rowname,
        y = Overall,
        label = sprintf("%0.1f", round(Overall, 1)) # sprintf ensures 1 decimal point
      ),
      hjust = "inward",
      colour = "black"
    ) +
    labs(title = title,
         x = "Term",
         y = "Scaled Variable Importance Measure") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.minor = element_blank())
}

# function to rename the categories of variables for better visualisation in graph
reformat_PREVENT1 <- function(VIM_df) {
  VIM_df <- VIM_df %>%
    mutate(rowname = as.factor(
      recode(
        rowname,
        "PROPERTYaccomm_tied_employment_armed_forces"	=	"PROPERTY Service tied accommodation",
        "PROPERTYfamily"	=	"PROPERTY Family"	,
        "PROPERTYhostel_B&B"	=	"PROPERTY Hostels & B&Bs",
        "PROPERTYroofless"	=	"PROPERTY Roofless"	,
        "PROPERTYfriends_partners"	=	"PROPERTY Friends & partners",
        "PROPERTYhome_ownership"	=	"PROPERTY Home ownership",
        "PROPERTYhospital"	=	"PROPERTY Hospital",
        "PROPERTYLA_tenancy"	=	"PROPERTY LA tenancy",
        "PROPERTYlodger"	=	"PROPERTY Lodger",
        "PROPERTYother"	=	"PROPERTY Other",
        "PROPERTYprison"	=	"PROPERTY Prison",
        "PROPERTYPRS_tenancy"	=	"PROPERTY PRS tenancy",
        "PROPERTYRSL_tenancy"	=	"PROPERTY RSL tenancy",
        "PROPERTYsofa_surfing"	=	"PROPERTY Sofa-surfing",
        "PROPERTYsupported_accom"	=	"PROPERTY Supported accommodation",
        "PROPERTYunknown_refused"	=	"PROPERTY Unknown or refused",
        "REASONasked_to_leave"	=	"REASON Asked to leave"	,
        "REASONdischarge_prison_hospital_institution"	=	"REASON Discharge"	,
        "REASONdispute_household_relationship_breakdown_non_violent"	=	"REASON Non-violent dispute"	,
        "REASONdispute_household_violent_abusive"	=	"REASON Violent abusive dispute"	,
        "REASONeviction_UK_borders_agency"	=	"REASON Eviction UK borders"	,
        "REASONfleeing_non_domestic_abuse_hate_crime_antisocial_behaviour"	=	"REASON Fleeing"	,
        "REASONgeneral_housing_options_advice"	=	"REASON HO advice"	,
        "REASONharrassment"	=	"REASON Harassment"	,
        "REASONother_financial_difficulties_benefit_issues"	=	"REASON Financial / benefits issues"	,
        "REASONother_landlord_action_resulting_termination"	=	"REASON Landlord action"	,
        "REASONother_reason_for_leaving_accom"	=	"REASON Other leaving accom"	,
        "REASONother_reason_loss_eviction_reposession"	=	"REASON Other loss accom"	,
        "REASONovercrowding"	=	"REASON Overcrowding"	,
        "REASONproperty_condition_or_accom_unsuitable"	=	"REASON Condition unsuitable"	,
        "REASONrisk_or_loss_service_tied_accommodation"	=	"REASON Service tied accom"	,
        "REASONtermination_tenancy_mortage_rent_arrears_default"	=	"REASON Rent arrears / default"	,
        "REASONapplicant_terminated_secure_accom"	=	"REASON Applicant termination"	,
        "financial_assistance_adviceTRUE"	=	"ACTIVITY Financial assistance"	,
        "client_informed_of_rights_under_homelessness_legislationTRUE"	=	"ACTIVITY Homelessness rights"	,
        "general_housing_advice_tenancy_rights_adviceTRUE"	=	"ACTIVITY HO advice"	,
        "referral_to_health_social_work_employment_servicesTRUE"	=	"ACTIVITY Referral services"	,
        "mediation_outreachTRUE"	=	"ACTIVITY Mediation"	,
        "help_to_move_to_other_propertyTRUE"	=	"ACTIVITY Help moving"	,
        "rent_repairs_referrals_neogtiation_with_landlordsTRUE"	=	"ACTIVITY Landlord negotiations"	,
        "help_to_remain_in_accommodationTRUE"	=	"ACTIVITY Help remaining"	,
        "other_activityTRUE"	=	"ACTIVITY Other"	,
        "tenancy_social_care_support_adaptations_to_propertyTRUE"	=	"ACTIVITY Social care & adaptations"	,
        "mortgage_homeownership_adviceTRUE"	=	"ACTIVITY Home ownership advice"	,
        "ethnicitywhite_scottish"	=	"ETHNICITY White Scottish"	,
        "ethnicitywhite_british"	=	"ETHNICITY White British" 	,
        "ethnicityother_white"	=	"ETHNICITY Other White"	,
        "ethnicitywhite_polish"	=	"ETHNICITY White Polish"	,
        "ethnicityasian"	=	"ETHNICITY Asian"	,
        "ethnicityother"	=	"ETHNICITY Other"	,
        "ethnicitynot_known"	=	"ETHNICITY Not known"	,
        "ethnicitycarribean_black"	=	"ETHNICITY Caribbean / Black"	,
        "ethnicitymixed"	=	"ETHNICITY Mixed"	,
        "ethnicityrefused"	=	"ETHNCITIY Refused",
        "activity_type_2TRUE"	=	"ACTIVITY TYPE 2"		,
        "activity_type_1TRUE"	=	"ACTIVITY TYPE 1"		,
        "activity_type_3TRUE"	=	"ACTIVITY TYPE 3"		,
        "activity_by_2TRUE"	=	"ACTIVITY BY 2"	,
        "activity_by_1TRUE"	=	"ACTIVITY BY 1"	,
        "activity_by_3TRUE"	=	"ACTIVITY BY 3"	,
        "RSNSCThave_to_leave_accommodation_household"	=	"RSNSCT Need to leave accom"	,
        "RSNSCTproperty_may_no_longer_be_suitable_available_in_future"	=	"RSNSCT Prop unavailable in future"	,
        "vul_count"	=	"VULNERABILITY COUNT"	,
        "mental_vulTRUE"	=	"VULNERABILITY Mental health"	,
        "lackofsupport_vulTRUE"	=	"VULNERABILITY Lack of support"	,
        "fin_vulTRUE"	=	"VULNERABILITY Financial"	,
        "phys_vulTRUE"	=	"VULNERABILITY Physical"	,
        "unmetneed_vulTRUE"	=	"VULNERABILITY Unmet need"	,
        "refused_vulTRUE" = "VULNERABILITY refused",
        "SUPPORT1" = "SUPPORT Yes",
        "SUPPORT2" = "SUPPORT Not known",
        "gender2" = "GENDER Female",
        "ageatapp" = "AGEATAPP",
        "previous_approachTRUE" = "Previous approach",
        "activity_count" = "ACTIVITY COUNT"
      )
    ))
}

# pulls out the top 5 most important variables
pull_top <- function(VIMdf, n) {
  VIMdf %>%
    mutate(Variable = forcats::fct_inorder(rowname),
           rowname = NULL) %>%
    head(5) %>% 
    pull(Variable) %>%
    as.character()
}

# set the destination filepath - should be in working directory then "/research/report/"
destfilepath <- "/homelessness-prevention-single-adults-scotland/research/report/"
```

Load in the data 

```{r}
# set the working directory for this chunk where the data is found 
setwd("/homelessness-prevention-single-adults-scotland/research/visualise/data/VIMs/")
folder <- list.files(pattern="*.csv")

# read the VIMs
VIMdfs <- lapply(folder, function(x) read_csv(x))

# use this naming convention for the graphs
names(VIMdfs) <- c("AdaboostM1", "Bagged Adaboost", "Elastic Net", "Logistic Regression", "Random Forest", "Extreme Gradient Boosting")

# rename the categories using the custom function built above
VIMdfs <- lapply(VIMdfs, reformat_PREVENT1)

# view the VIM data frames
VIMdfs
```

Graph the individual VIMs

```{r}
# graph the top 20 VIMs for each algorithm putting the full algorithm name in each chart title
VIM_graphs <- lapply(1:length(VIMdfs), 
                     function(x) 
                       plotvim(VIMdfs[[x]], 
                               title = paste(names(VIMdfs[x]))))

# name the individual graphs in the list of graphs - used for saving them out
names(VIM_graphs) <- c("Ada", "BAda", "ENet", "LR", "RF", "XGB")

# view the graphs
VIM_graphs

# save each individual graph 
lapply(names(VIM_graphs),
       function(x)
         ggsave(
           filename = file.path(destfilepath, 
                                paste(x, 
                                      "_VIM.png",
                                      sep = "")),
           plot = VIM_graphs[[x]],
           width = 5,
           height = 3.5
         ))
```

